---
title: 'REPORT: A Research of the Effects of Disaster Evnets on Public Health and Wealth in US by Means of Analysis of NOAA Disaster Event Data Base' 
output: html_document
---

## Synopsis ##
In this report the author aims to describe the process of analysis of a dataset of disastrous event observations in order to establish the event types that are most damaging to public wealth and dangerous for public health. To accomplish this task author attempts to explore the NOAA Storm Database, analyse its contents and structure and extract the information related to the subject of interest. 

## 1 Data Processing ##

### 1.1 Loading the Raw Data ###
Data was taken from: [link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]
The following information is contained within the dataset

```{r}
obj <- read.csv("repdata_data_StormData.csv.bz2", stringsAsFactors = F)
```

### 1.2 Exploring the Dataset###
Some exploration of the data needs to be made first to get some more details on the dataset. These details can prove useful when considering further dataset alterations.

#### 1.2.1 Dimensions: ####
```{r}
dim(obj)
```

#### 1.2.2 Nature of data: ####
```{r}
str(obj)
```

#### 1.2.3 Simplification: ####
```{r}
names(obj)[1] <- "STATE"            # removes underscores
names(obj) <- tolower(names(obj))   # lower case for ease of typing
obj$evtype <- tolower(obj$evtype)   # lower case to simplify aggregation
obj$date <- as.Date(obj$bgn_date, "%m/%d/%Y") # Date conversion
```

#### 1.2.4 Unique observations: ####
```{r}
uniqueD <- data.frame(matrix(NA, nrow=1, ncol = (length(names(obj)))))
names(uniqueD) <- names(obj)

for (i in 1:length(names(obj))) {
    uniqueD[1, i] <- length(unique(obj[, i]))
}
uniqueD
```

Due to so many different events being observed and a rather long observation period it is important to establish the observation pattern, as this can seriously affect any aggregations in case there is a significant bias in event observation. For instance an important and dangerous event may have just been started being observed few years ago, but a moderately dangerous event which was observed for decades may appear much more damaging if damage absolute totals are compared.

```{r}
x3 <- aggregate(as.Date(bgn_date, "%m/%d/%Y") ~ evtype, min, data = obj)
x3 <- x3[order(x3[,2], decreasing=F),]
head(x3)
```


### 1.3 Exploration Conclusions###
Turns out that period of observations 1950 - 1993 mostly contains observations of just three disaster types, while 1993 is the first year of a more diverse information availability. Additionally it is possible that the data available on the pre-1993 observation describe only the most destructive events which again may add bias to the analysis

However, a significant part of event types observed since 1993 are one time only and a brief overview of them shows lack of standardisation in the observation records.

A lot of remaining event types are variations of a much narrower event set.

Additional important facts revealed:

- Coordinates are stored in the XXXX format instead of XX.XX
- Latitude values which were supposed to be negative are stored as positive numbers
- Property (propdmgexp) and Crop damage (cropdmgexp) magnitude factors include much more different values than "h ", "k", "m" and "b" described in the _"NOAA Storm Data preparation"_ document
- Dataset contains a total of 898 different event types as opposed to expected 48 types outlined in the _"NOAA Storm Data preparation"_ document
- All items contained in "Refnum" column are unique, which means that each row describes a different event

### 1.4 Post-Exploration Action Plan###
The following tweaking plan is devised:

- Removing the 1950-1993 data as it brings bias to the review
- Event types to be reclassified to satisfy the NOAA Classification
- Only use damage magnitude observations mentioned in the _"NOAA Storm Data preparation"_ document
- The analysis is aimed at presenting the top 10 of most dangerous/damaging events

### 1.5 Data Preparation ###
Coordinates correction:
```{r}
mydata<- cbind(obj$latitude,obj$longitude)
mydata<-as.data.frame(mydata)
mydata <-cbind(mydata, obj$evtype)
colnames(mydata) <- c("lat", "long", "evtype")
mydata[, 1] <- mydata[, 1]/100
mydata[, 2] <- mydata[, 2]/(-100)
```

Addition of damage and fatalities data:
```{r}
mydata <- cbind(mydata, obj[, c(23:28)]) # adds in damage and fatalities
mydata <- cbind(mydata, as.Date(obj[, 2], "%m/%d/%Y"))
colnames(mydata)[10] <- "date" 
```

Filtering out the pre-1993 records:
```{r}
mydataC <- subset(mydata, mydata$date > "1992-12-31")
```

Reclassification of events:
The event types will be reclassified basing on the list included in the _"NOAA Storm Data preparation"_ document. This list consists of 48 different event types. The author aims at maping majority of the event types included in the dataset to the NOAA types. The following issues were considered:
- Unclassifiable/Hardly classifiable event types will not be reclassified and will be missed out
- Events where more than one classification is possible will be classified as belonging to both (or more where applicable) NOAA event types in order to retain data consistence and relevance
- Reclassification dictionary (presented below) is compiled by the author manually, through analysis of the available unique event types. This method of sorting was employed by the auther as a significant number of the events were unclassiyable using the regular expressions. The possibility of employing Natural Language Processing was considered, but author did not magae to attain satisfactory results using this approach. Hence the manual mapping which is believe to be reasonably accurate, with multi-classification helping in ambiguous cases.
```{r}
#----Names----
et1 <- "astronomical low tide"              # Astronomical Low Tide
et2 <- "avalanche"                          # Avalanche 
et3 <- c("blizzard",
         "blizzard and extreme wind chil",
         "blizzard/heavy snow",
         "ground blizzard",
         "heavy snow/blizzard",
         "high wind/blizzard")              # Blizzard 
et4 <- c("coastal flood",
         "coastal flooding",
         "coastal flooding/erosion",
         "coastal/tidal flood",
         "beach flood",
         "erosion/cstl flood",
         "tidal flooding",
         "cstl flooding/erosion")           # Coastal flood
et5 <- c("cold",
         "cold temperature",
         "cold wave",
         "cold wind chill temperatures",
         "high winds/cold",
         "snow and cold",
         "cold/wind chill",
         "cold and frost",
         "cold temperatures",
         "cold weather",
         "wind chill",
         "snow/cold")                       # Cold/Wind Chill
et6 <- NA                                   # Debris Flow
et7 <- c("dense fog", 
         "fog", 
         "patchy dense fog")                # Dense Fog
et8 <- c("dense smoke", 
         "smoke")                           # Dense Smoke
et9 <- c("drought",
         "drought/excessive heat",
         "snow drought")                    # Drought 
et10 <- "dust devil"                        # Dust Devil
et11 <- c("blowing dust", 
          "dust storm")                     # Dust Storm
et12 <- c("drought/excessive heat",
          "record heat",
          "excessive heat",
          "record/excessive heat",
          "extreme heat")                   # Excessive Heat
et13 <- c("excessive cold", 
          "extreme cold/wind chill",
          "record cold",
          "unseasonably cold",
          "extreme wind chill",
          "extreme windchill temperatures",
          "bitter wind chill temperatures",
          "cold/wind chill",
          "extreme cold",
          "extreme/record cold",
          "record cold/frost",
          "unusually cold",
          "extreme cold/wind chill",
          "extreme windchill",
          "cold wind chill temperatures",
          "wind chill")                     # Extreme Cold/Wind Chill
et14 <- c("flash flood",
          "flash flood/ flood",
          "flash flooding/flood",
          "flood/flash",
          "flash flood - heavy rain",
          "flash flood/flood",
          "flash floods",
          "flood/flash flood",
          "flash flood from ice jams",
          "flash flooding",
          "flood flash",
          "flood/flash flooding")           # Flash Flood
et15 <- c("urban/street flooding",
          "urban/small stream flooding",
          "urban/small stream flood",
          "urban/small stream  flood",
          "urban small stream flood",
          "urban floods",
          "urban flooding",
          "urban flood",
          "urban and small stream floodin",
          "urban and small stream flood",
          "thunderstorm winds/ flood",
          "street flooding",
          "street flood",
          "snowmelt flooding",
          "small stream/urban flood",
          "small stream flooding",
          "small stream flood",
          "small stream and urban flood",
          "rural flood",
          "river flooding",
          "river flood",
          "river and stream flood",
          "minor flooding",
          "major flood",
          "ice jam flooding",
          "heavy rains/flooding",
          "heavy rain/flooding",
          "floods",
          "flooding",
          "flood/flash flooding",
          "flood",
          "flood & heavy rain")             # Flood 
et16 <- c("frost",
          "frost/freeze",
          "snow/sleet/freezing rain",
          "snow/freezing rain",
          "snow freezing rain",
          "sleet/freezing rain",
          "light freezing rain",
          "heavy snow/freezing",
          "hard freeze",
          "frost/freeze",
          "freezing rain/snow",
          "freezing rain/sleet",
          "freezing rain",
          "freezing rain and sleet",
          "freezing drizzle",
          "freeze",
          "damaging freeze",
          "agricultural freeze")            # Frost/Freeze
et17 <- c("cold air funnel",
          "funnel",
          "funnel clouds",
          "cold air funnels",
          "funnel cloud",
          "thunderstorm winds funnel clou") # Funnel Cloud
et18 <- c("freezing fog",
          "ice fog")                        # Freezing Fog
et19 <- c("hail",
          "hail 1.75",
          "hail 175",
          "hail 80",
          "hail/winds",
          "non severe hail",
          "thunderstorm winds/hail",
          "hail 0.75",
          "hail 100",
          "hail 275",
          "hail damage",
          "hailstorm",
          "small hail",
          "tstm wind/hail",
          "hail 1.00",
          "hail 150",
          "hail 75",
          "hail/wind",
          "marine hail",
          "thunderstorm winds hail")        # Hail
et20 <- c("heat",
          "heat wave",
          "heat waves")                     # Heat
et21 <- c("excessive rain",
          "unseasonal rain",
          "tstm heavy rain",
          "snow/sleet/freezing rain",
          "snow/freezing rain",
          "snow freezing rain",
          "sleet/freezing rain",
          "record rainfall",
          "record low rainfall",
          "rain/snow",
          "rain",
          "prolonged rain",
          "monthly rainfall",
          "light freezing rain",
          "hvy rain",
          "heavy rains/flooding",
          "heavy rains",
          "heavy rainfall",
          "heavy rain/wind",
          "heavy rain/severe weather",
          "heavy rain/flooding",
          "heavy rain",
          "heavy rain and wind",
          "flood & heavy rain",
          "flash flood - heavy rain",
          "excessive rainfall")             # Heavy Rain
et22 <- c("snow squalls",
          "snow squall",
          "snow and heavy snow",
          "record winter snow",
          "record snowfall",
          "record snow",
          "high wind/heavy snow",
          "heavy snow-squalls",
          "heavy snow/squalls",
          "heavy snow/ice",
          "heavy snow/ice storm",
          "heavy snow/freezing rain",
          "heavy snow/blizzard",
          "heavy snow",
          "heavy snow squalls",
          "heavy snow and ice",
          "heavy snow and ice storm",
          "heavy snow and high winds",
          "heavy lake snow",
          "excessive snow",
          "blizzard/heavy snow",
          "accumulated snowfall")           # Heavy Snow
et23 <- c( "heavy surf",
           "heavy surf/high surf",
           "high surf", 
           "high surf advisory", 
           "rip currents/heavy surf", 
           "rough surf")                    # High Surf
et24 <- c("gusty thunderstorm winds",
          "gusty winds",
          "heavy rain/wind",
          "high wind",
          "high wind and high tides",
          "high wind/blizzard",
          "high winds",
          "high winds 66",
          "high winds/cold",
          "severe thunderstorm winds",
          "strong wind",
          "strong winds",
          "whirlwind",
          "gusty wind",
          "hail/wind",
          "heavy snow and high winds",
          "high wind (g40)",
          "high wind damage",
          "high wind/heavy snow",
          "high winds 63",
          "high winds 80",
          "high winds/snow",
          "snow/high winds",
          "strong wind gust",
          "thuderstorm winds",
          "wind",
          "wind gusts")                     # High Wind
et25 <- c("hurricane",
          "hurricane erin",
          "hurricane/typhoon",
          "hurricane-generated swells",
          "hurricane felix",
          "hurricane edouard",
          "hurricane opal",
          "typhoon")                        # Hurricane (Typhoon)
et26 <- c("falling snow/ice", 
          "heavy snow and ice",
          "heavy snow and ice storm",
          "heavy snow/ice",
          "heavy snow/ice storm",
          "ice storm",
          "snow/ice storm")                 # Ice Storm
et27 <- c("heavy lake snow",
          "lake-effect snow",
          "lake effect snow" )              # Lake-Effect Snow
et28 <- "lakeshore flood"                   # Lakeshore Flood
et29 <- c( "lightning",
           "thunderstorm winds lightning")  # Lightning
et30 <- "marine hail"                       # Marine Hail
et31 <- "marine high wind"                  # Marine High Wind
et32 <- "marine strong wind"                # Marine Strong Wind
et33 <- c("marine thunderstorm wind",
          "marine tstm wind")               # Marine Thunderstorm Wind
et34 <- c("rip current",
          "rip currents",
          "rip currents/heavy surf")        # Rip Current
et35 <- "seiche"                            # Seiche
et36 <- c("coastal storm",
          "storm surge",
          "storm surge/tide",
          "coastal surge")                  # Storm Surge/Tide
et37 <- c( "strong wind",
           "strong wind gust",
           "strong winds",
           "gusty wind",
           "gusty winds",
           "non tstm wind",
           "wind",
           "wind advisory",
           "wind damage",
           "wind gusts",
           "winds")                         # Strong Wind
et38 <- c("gusty thunderstorm wind",
          "tstm winds",
          "tstm wind/hail",
          "tstm wind",
          "tstm wind 55",
          "tstm wind 52",
          "tstm wind 51",
          "tstm wind (g45)",
          "tstm wind (g40)",
          "tstm heavy rain",
          "thundertorm winds",
          "thunderstrom winds",
          "thunderstormw winds",
          "thunderstorms",
          "thunderstorms winds",
          "thunderstorms wind",
          "thunderstorm",
          "thunderstorm windss",
          "thunderstorm winds/hail",
          "thunderstorm winds/ flood",
          "thunderstorm winds.",
          "thunderstorm winds",
          "thunderstorm winds lightning",
          "thunderstorm winds hail",
          "thunderstorm winds g",
          "thunderstorm winds funnel clou",
          "thunderstorm winds and",
          "thunderstorm wind/ trees",
          "thunderstorm wind",
          "thunderstorm wind g60",
          "thunderstorm wind g52",
          "thunderstorm wind g50",
          "thunderstorm wind 60 mph",
          "thunderstorm wind 50",
          "thunderstorm damage",
          "thunderstorm  winds",
          "severe thunderstorms",
          "severe thunderstorm",
          "severe thunderstorm winds",
          "marine tstm wind",
          "marine thunderstorm wind",
          "gusty thunderstorm winds",
          " tstm wind")                     # Thunderstorm Wind
et39 <- c("tornado",
          "waterspout-tornado",
          "waterspout/tornado",
          "waterspout/ tornado",
          "tornadoes",
          "tornado f3",
          "tornado f2",
          "tornado f1",
          "tornado f0")                     # Tornado
et40 <- "tropical depression"               # Tropical Depression
et41 <- c("tropical depression",
          "tropical storm",
          "tropical storm dean",
          "tropical storm jerry")           # Tropical Storm
et42 <- "tsunami"
et43 <- c("volcanic ash",
          "volcanic ashfall")               # Volcanic Ash
et44 <- c("waterspout",
          "waterspout-",
          "waterspout-tornado",
          "waterspout/ tornado",
          "waterspout/tornado",
          "waterspouts")                    # Waterspout 
et45 <- c("wild fires",
          "wild/forest fire",
          "wildfire",
          "wildfires",
          "brush fire")                     # Wildfire
et46 <- c("winter storm",
          "winter storms")                  # Winter Storm
et47 <- c("winter mix",
          "winter weather",
          "winter weather mix",
          "winter weather/mix",
          "wintery mix",
          "wintry mix")                     # Winter Weather
et48 <- c("freezing rain and sleet",
          "freezing rain/sleet",
          "light snow and sleet",
          "sleet",
          "sleet storm",
          "sleet/freezing rain",
          "sleet/snow",
          "snow and sleet",
          "snow/sleet",
          "snow/sleet/freezing rain")       # Sleet
#----Names-End----

# List containing vectors with event types from the dataset
# to be used in reclassification loop
EvClass <- list(et1, et2, et3, et4, et5, et6, et7, et8, et9, et10, et11, 
                et12, et13, et14, et15, et16, et17, et18, et19, et20, et21, 
                et22, et23, et24, et25, et26, et27, et28, et29, et30, et31,
                et32, et33, et34, et35, et36, et37, et38, et39, et40, et41, 
                et42, et43, et44, et45, et46, et47, et48)

# Vector containing NOAA event types - to be used in reclassification loop
EvName <- c("Astronomical_Low_Tide", "Avalanche", "Blizzard", 
               "Coastal_Flood", "Cold_OR_Wind_Chill", "Debris_Flow", 
               "Dense_Fog", "Dense_Smoke", "Drought", "Dust_Devil", 
               "Dust_Storm", "Excessive_Heat", "Extreme_Cold_OR_Wind_Chill",
               "Flash_Flood", "Flood", "Frost_OR_Freeze","Funnel_Cloud",
               "Freezing_Fog", "Hail", "Heat", "Heavy_Rain", "Heavy_Snow",
               "High_Surf", "High_Wind", "Hurricane_Typhoon", "Ice_Storm",
               "LakeEffect_Snow", "Lakeshore_Flood", "Lightning", 
               "Marine_Hail", "Marine_High_Wind", "Marine_Strong_Wind", 
               "Marine_Thunderstorm_Wind", "Rip_Current", "Seiche", 
               "Storm_Surge_OR_Tide", "Strong_Wind", "Thunderstorm_Wind",
               "Tornado", "Tropical_Depression", "Tropical_Storm", "Tsunami",
               "Volcanic_Ash", "Waterspout", "Wildfire", "Winter_Storm",
               "Winter_Weather", "Sleet")

# Reclassification()
for (i in 1:length(EvName)){
    mydataC[mydataC$evtype %in% EvClass[[i]], 10+i] <- EvName[i]
}
colnames(mydataC)[11:58] <- EvName                     
```

## 2 Results ##
This section aims to provide analysis for the two following problems:

- Across the United States, which types of events are most harmful with respect to population health?
- Across the United States, which types of events have the greatest economic consequences?

### 2.1 Results: Public Health ###

Preparation
```{r}
# empty DFs to be used later in preparation of individual subsets
PHDangers <- data.frame()
PWDangersP <- data.frame()
PWDangersC <- data.frame()
```

Subsetting of PH events
```{r}
for (i in 1:length(EvName)){
    if (i != 6) {
        tdf <- aggregate(cbind(fatalities, injuries) ~ mydataC[,i+10], sum, data = mydataC)
        names(tdf) <- c("evtype", "fatalities", "injuries")
        tdf$evcount <- length(subset(mydataC[,i+10], 
                                     mydataC [,i+10] != "NA"))
        PHDangers <- rbind(PHDangers, tdf)
        
    }
}
fatTop2 <- PHDangers
```

Subsetting top ten most dangerous event types in terms of fatalities
```{r}
# Sum
fatTopF <- fatTop2[order(fatTop2[,2], decreasing=F),]
fatTopF <- tail(fatTopF, 10)
fatTopF

# Mean
fatTopFM <- transform(fatTop2, new = round(fatalities / evcount, digits = 0))
fatTopFM <- fatTopFM[order(fatTopFM[,5], decreasing=F),]
fatTopFM <- tail(fatTopFM, 10)

```

Subsetting top ten most dangerous event types in terms of injuries
```{r}
# Sum 
fatTopI <- fatTop2[order(fatTop2[,3], decreasing=F),]
fatTopI <- tail(fatTopI, 10)
fatTopI

# Mean
fatTopIM <- transform(fatTop2, new = round(injuries / evcount, digits = 0))
fatTopIM <- fatTopIM[order(fatTopIM[,5], decreasing=F),]
fatTopIM <- tail(fatTopIM, 10)
```

Public Health dangers - plot
Note - Mean plots, represent mean values of the top 10 most dangeous events identified by the comparison of Total values and not the top mean values across all data set. This is due to the fact that there are many one time events which could get into the plot, but would hardly be relevant to our task.
```{r,  fig.height = 12, fig.width = 10}
par(mfrow=c(2,2))
par(las=2)
par(mar=c(5,8,2,2))
par(oma=c(0,0,3,0))

# sum plots
barplot(fatTopF$fatalities, main="Total Fatalities", horiz=TRUE,
        names.arg=fatTopF$evtype, cex.names=0.8, cex.main = 1)
barplot(fatTopI$injuries, main="Total Injuries", horiz=TRUE,
        names.arg=fatTopI$evtype, cex.names=0.8, cex.main = 1)

# mean plots
barplot(fatTopFM$new, main="Mean Fatalities", horiz=TRUE,
        names.arg=fatTopFM$evtype, cex.names=0.8, cex.main = 1)
barplot(fatTopIM$new, main="Mean Injuries", horiz=TRUE,
        names.arg=fatTopIM$evtype, cex.names=0.8, cex.main = 1)

# main caption
title("Public Health Dangers", outer=T)
```

### 2.2 Results: Public Wealth ###
Subsetting the public wealth dangers
```{r}
mydataL <- mydataC
mydataL [, 7] <- as.factor(tolower(mydataL [, 7]))
mydataL [, 9] <- as.factor(tolower(mydataL [, 9]))
mydataPW2Prop <- subset(mydataL, 
                        propdmgexp %in% c("k","m","b"))
mydataPW2Crop <- subset(mydataL, 
                        cropdmgexp %in% c("k","m","b"))

# multiplication of property damage
for (i in 1:length(mydataPW2Prop)) {
    if (mydataPW2Prop[i, 6] == "k") {
        mydataPW2Prop[i, 6] <- mydataPW2Prop[i, 6]*1000
    } else if (mydataPW2Prop[i, 6] == "m") {
        mydataPW2Prop[i, 6] <- mydataPW2Prop[i, 6]*1000000
    } else {
        mydataPW2Prop[i, 6] <- mydataPW2Prop[i, 6]*1000000000
    }
}

# multiplication of crop damage
for (i in 1:length(mydataPW2Crop)) {
    if (mydataPW2Crop[i, 8] == "k") {
        mydataPW2Crop[i, 8] <- mydataPW2Crop[i, 8]*1000
    } else if (mydataPW2Crop[i, 8] == "m") {
        mydataPW2Crop[i, 8] <- mydataPW2Crop[i, 8]*1000000
    } else {
        mydataPW2Crop[i, 8] <- mydataPW2Crop[i, 8]*1000000000
    }
}

```

The crop damage table
```{r}
# i <- 1
for (i in 1:length(EvName)){
    if (i != 6) {
        tdf2 <- aggregate(cropdmg ~ mydataPW2Crop[,i+10], sum, 
                          data = mydataPW2Crop)
        names(tdf2) <- c("evtype", "cropdmg")
        tdf2$evcount <- length(subset(mydataPW2Crop[,i+10], 
                                      mydataPW2Crop[,i+10] != "NA"))
        PWDangersC <- rbind(PWDangersC, tdf2)        
    }
}
cropTop2 <- PWDangersC
```

The property damage table
```{r}
for (i in 1:length(EvName)){
    if (i != 6) {
        tdf3 <- aggregate(cropdmg ~ mydataPW2Prop[,i+10], sum, 
                          data = mydataPW2Prop)
        names(tdf3) <- c("evtype", "propdmg")
        tdf3$evcount <- length(subset(mydataPW2Prop[,i+10], 
                                      mydataPW2Prop[,i+10] != "NA"))
        PWDangersP <- rbind(PWDangersP, tdf3)        
    }
}
propTop2 <- PWDangersP
```

Subsetting top ten most dangerous event types in terms of property damage
```{r}
# sum
propTopN <- propTop2[order(propTop2[,2], decreasing=F),]
propTopN <- tail(propTopN, 10)
propTopN

# mean 
propTopNM <- transform(propTop2, new = round(propdmg / evcount, digits = 0))
propTopNM <- propTopNM[order(propTopNM[,4], decreasing=F),]
propTopNM <- tail(propTopNM, 10)
propTopNM
```

Subsetting top ten most dangerous event types in terms of crop damage
```{r}
# sum
cropTopN <- cropTop2[order(cropTop2[,2], decreasing=F),]
cropTopN <- tail(cropTopN, 10)
cropTopN

# mean
cropTopNM <- transform(cropTop2, new = round(cropdmg / evcount, digits = 0))
cropTopNM <- cropTopNM[order(cropTopNM[,4], decreasing=F),]
cropTopNM <- tail(cropTopNM, 10)
cropTopNM
```

Public Wealth dangers - Plot
Note - Mean plots, represent mean values of the top 10 most dangeous events identified by the comparison of Total values and not the top mean values across all data set. This is due to the fact that there are many one time events which could get into the plot, but would hardly be relevant to our task.

```{r,  fig.height = 12, fig.width = 10}
par(mfrow=c(2,2))
par(las=2)
par(mar=c(5,8,4,2))
par(oma=c(0,0,3,0))

#sum
barplot(propTopN$propdmg, main="Total Property Damage", horiz=TRUE,
        names.arg=propTopN$evtype, cex.names=0.8, cex.main = 1)
barplot(cropTopN$cropdmg, main="Total Crop Damage", horiz=TRUE,
        names.arg=cropTopN$evtype, cex.names=0.8, cex.main = 1)

#mean
barplot(propTopNM$new, main="Mean Property Damage", horiz=TRUE,
        names.arg=propTopNM$evtype, cex.names=0.8, cex.main = 1)
barplot(cropTopNM$new, main="Mean Crop Damage", horiz=TRUE,
        names.arg=cropTopNM$evtype, cex.names=0.8, cex.main = 1)

# main caption
title("Public Wealth Dangers", outer=T)
```